---
title: "EDA I.4: Introduction to R and Data Analysis"
subtitle: "Analyse de Données Master I, MFA et MIDS"
author: "Stéphane Boucheron"
institute: "Université de Paris"
date: "2021/12/21 (updated: `r Sys.Date()`)"
params:
  title: "Beyond dplyr"
  curriculum: "Master I MIDS & MFA"
  coursetitle: "Analyse Exploratoire de Données"
  lecturer: "Stéphane Boucheron"
  homepage: "http://stephane-v-boucheron.fr/courses/eda/"
  curriculum_homepage: "https://master.math.univ-paris-diderot.fr/annee/m1-mi/"
  institution: "Université de Paris"
output:
  xaringan::moon_reader:
    css: ["header-footer.css", "xaringan-themer.css"]
    lib_dir: libs
    seal: false
    includes:
      in_header:
        - 'toc.html'
    nature:
      nature:
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
bibliography: mon_chapeau.bib
---
name: inter-slide
class: left, middle, inverse

{{ content }}

---
name: layout-general
layout: true
class: left, middle

```{r setup, child="loaders_fixers.Rmd", echo=FALSE, message=FALSE, warning=FALSE}

```


---


```{r child='title_slide.Rmd'}

```


---
template: inter-slide

## `r fontawesome::fa("map", fill="white")`

### [DBI](#DBI)

### [sqldf](#sqldf)

### [data.table](#data.table)

### [dtplyr](#dtplyr)

---
template: inter-slide
name: DBI


## `DBI`


---
template: inter-slide
name: sqldf

## `sqldf`


---
template: inter-slide
name: datatable

## `datatable`


---

### Package `datatable`


```r
pacman::p_load("data.table")
```

> Data manipulation operations such as `subset`, `group`, `update`, `join` etc., are all inherently related

> Keeping these related operations together allows for:

--

> - concise and consistent syntax ...

--

> - performing analysis without the burden of mapping each operation to a particular function from a huge set

--

> - automatically optimizing operations internally, and very effectively, ..., leading to very fast and memory efficient code.

???

> Briefly, if you are interested in reducing programming and compute time tremendously, then this package is for you.


---


```{r}
dtflights <-  data.table(flights)

dtflights |>
  head(20) |>
  DT::datatable()
```

---


```{r}
dtflights[
  dest %in% c('ATL', 'LAX') & origin == 'JFK', #<< filtering
  .(dest, carrier, time_hour, retard= arr_delay - dep_delay), #<< projecting
  ] |>
  head(20) |>
  DT::datatable()
```

???

> Within the frame of a `data.table`, columns can be referred to as if they are variables.

> We  refer to `origin` and `dest` as if they were variables.

> In the `j` expression, we can compute new columns ...


---


### Aggregation with `datatable`

.fl.w-50.pa2[

```{r}
dtflights[
  ,               #&lt;&lt; (no) filtering
  .(n=.N),        #&lt;&lt; summarize
  by = .(carrier) #&lt;&lt; aggregation
  ] |&gt;
  DT::datatable()
```

<div id="htmlwidget-6358a46d1399716e8228" style="width:100%;height:auto;" class="datatables html-widget"></div>
<script type="application/json" data-for="htmlwidget-6358a46d1399716e8228">{"x":{"filter":"none","vertical":false,"data":[["1","2","3","4","5","6","7","8","9","10","11","12","13","14","15","16"],["UA","AA","B6","DL","EV","MQ","US","WN","VX","FL","AS","9E","F9","HA","YV","OO"],[58665,32729,54635,48110,54173,26397,20536,12275,5162,3260,714,18460,685,342,601,32]],"container":"<table class=\"display\">\n  <thead>\n    <tr>\n      <th> <\/th>\n      <th>carrier<\/th>\n      <th>n<\/th>\n    <\/tr>\n  <\/thead>\n<\/table>","options":{"columnDefs":[{"className":"dt-right","targets":2},{"orderable":false,"targets":0}],"order":[],"autoWidth":false,"orderClasses":false}},"evals":[],"jsHooks":[]}</script>
]


.fl.w-50.pa2[


```sql
SELECT carrier, COUNT(*) AS n
FROM flights
GROUP BY carrier
ORDER BY n DESC
```

]

---
template: inter-slide
name: dtplyr

## `dtplyr`

---

`dtplyr`: for large, in-memory datasets.

 Translates your `dplyr` code to high performance `data.table` code



---
template: inter-slide
name: dbplyr

## `dbplyr`

---

`dbplyr`: for data stored in a relational database. Translates your `dplyr` code to `SQL`


---

```{r child='closing_slide.Rmd'}

```
