---
title: "EDA XIV: Tidy models III"
subtitle: "Analyse de Données Master I, MFA et MIDS"
author: "Stéphane Boucheron"
institute: "Université de Paris"
date: "2021/12/21 (updated: `r Sys.Date()`)"
params:
  title: "Introduction to tidy modeling III"
  curriculum: "Master I MIDS & MFA"
  coursetitle: "Analyse Exploratoire de Données"
  lecturer: "Stéphane Boucheron"
  homepage: "http://stephane-v-boucheron.fr/courses/eda/"
  curriculum_homepage: "https://master.math.univ-paris-diderot.fr/annee/m1-mi/"
  institution: "Université de Paris"
output:
  xaringan::moon_reader:
    css: ["header-footer.css", "xaringan-themer.css"]
    lib_dir: libs
    seal: false
    includes:
      in_header:
        - 'toc.html'
    nature:
      nature:
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
bibliography: mon_chapeau.bib
---
name: inter-slide
class: left, middle, inverse

{{ content }}

---
name: layout-general
layout: true
class: left, middle

```{r setup, child="loaders_fixers.Rmd", echo=FALSE, message=FALSE, warning=FALSE}

```

---


```{r child="title_slide.Rmd"}

```

---
template: inter-slide

## `r fontawesome::fa("map", fill="white")`

## [Introduction]()
## [Data splitting]()
## [Modeling]()
## [Estimating performance]()
## [Resampling to the rescue]()
## [Fit a model with resampling]()
## [Session information]()

---
template: inter-slide

## Introduction


???


---

```{r}
knitr::include_url("https://www.tidymodels.org/books/moderndive/")
```

---

> we have built a model and preprocessed data with a `recipe`. We also introduced `workflows` as a way to bundle a `parsnip` model and `recipe` together. Once we have a model trained, we need a way to measure how well that model predicts new data. This tutorial explains how to characterize model performance based on resampling statistics.


```{r}
library(tidymodels) # for the rsample package, along with the rest of tidymodels

# Helper packages
library(modeldata)  # for the cells data
```


---

```{r}
data(cells, package = "modeldata")
cells
#> # A tibble: 2,019 x 58
#>   case  class angle_ch_1 area_ch_1 avg_inten_ch_1 avg_inten_ch_2 avg_inten_ch_3
#>   <fct> <fct>      <dbl>     <int>          <dbl>          <dbl>          <dbl>
#> 1 Test  PS        143.         185           15.7           4.95           9.55
#> 2 Train PS        134.         819           31.9         207.            69.9
#> 3 Train WS        107.         431           28.0         116.            63.9
#> 4 Train PS         69.2        298           19.5         102.            28.2
#> 5 Test  PS          2.89       285           24.3         112.            20.5
#> # … with 2,014 more rows, and 51 more variables: avg_inten_ch_4 <dbl>,
#> #   convex_hull_area_ratio_ch_1 <dbl>, convex_hull_perim_ratio_ch_1 <dbl>,
#> #   diff_inten_density_ch_1 <dbl>, diff_inten_density_ch_3 <dbl>, …

```
---

```{r child='closing_slide.Rmd'}

```
