---
title: "EDA I: Introduction to R and Data Analysis"
subtitle: "Analyse de Données Master I, MFA et MIDS"
author: "Stéphane Boucheron"
institute: "Université de Paris"
date: "2021/12/21 (updated: `r Sys.Date()`)"
params:
  title: "Introduction to visualization"
  curriculum: "Master I MIDS & MFA"
  coursetitle: "Analyse Exploratoire de Données"
  lecturer: "Stéphane Boucheron"
  homepage: "http://stephane-v-boucheron.fr/courses/eda/"
  curriculum_homepage: "https://master.math.univ-paris-diderot.fr/annee/m1-mi/"
  institution: "Université de Paris"
output:
  xaringan::moon_reader:
    css: ["header-footer.css", "xaringan-themer.css"]
    lib_dir: libs
    seal: false
    includes:
      in_header:
        - 'toc.html'
    nature:
      nature:
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
bibliography: mon_chapeau.bib
---
name: inter-slide
class: left, middle, inverse

{{ content }}

---
name: layout-general
layout: true
class: left, middle

```{r setup, child="loaders_fixers.Rmd", echo=FALSE, message=FALSE, warning=FALSE}

```

```{r, echo=FALSE}
#xaringanExtra::use_scribble()
xaringanExtra::use_webcam()
```

---


```{r child='title_slide.Rmd'}

```


---
template: inter-slide

## `r fontawesome::fa("map", fill="white")`

### [Grammar of Graphics illustrated](#gg)

### [Do It Yourself](#diy)

### [Books](#books)

---
template: inter-slide

## Grammar of Graphics

---

We will use the _Grammar of Graphics_ approach to visualization 

The expression _Grammar of Graphics_ was coined by Leiland Wilkinson to describe a principled approach to visualization in Data Analysis (EDA)

A plot is organized around data (a table with rows (observations) and columns (variables)) 

A plot is a graphical object that can be built layer by layer  

Building a graphical object consists in chaining elementary operations 

The acclaimed TED presentation by Hans Rossling illustrates the Grammar of Graphics approach in a very appealing way

---
### Grammar of Graphics in Action

```{r rossling, eval=TRUE}
knitr::include_url("https://www.youtube.com/embed/jbkSRLYSojo")
```

???

---

### [`r fontawesome::fa("link")` Gapminder](https://www.gapminder.org)

[Gapminder](https://www.gapminder.org) is a Swedish foundation. Its website states

> Gapminder fights devastating misconceptions about global development. Gapminder produces free teaching resources making the world understandable based on reliable statistics. Gapminder promotes a fact-based worldview everyone can understand.

Hans Rosling was a founder of Gapminder

---
template: inter-slide
name: diy

## Do It Yourself


---

We will reproduce the animated demonstration using 

- `ggplot2` an implementation of grammar of graphics in `r fontawesome::fa("r-project")`

- `plotly` a bridge between `r fontawesome::fa("r-project")` and the javascript library `D3.js`



---
### `r fontawesome::fa("tools")` Install and load packages

.fl.w-50.pa2.f6[

The `gapminder` data can be loaded from pakage `gapminder`

- `r fontawesome::fa("download")` Packages can be _installed_ from `CRAN` using `install.packages()`

- `r fontawesome::fa("hdd")` Once a package is installed on the hard drive, it has to be _loaded_ in the current session using `require()` or `library()`

- `r fontawesome::fa("hdd")` Function `p_load` from package `pacman` installs the package if needed and loads it


]

.fl.w-50.pa2[

```{r load_gapminder_again, echo=TRUE}
pacman::p_load("tidyverse")
pacman::p_load("gapminder")
```

- `r fontawesome::fa("magic")` Loading `tidyverse` loads a collection of packages dedicated to table manipulations and graphics including `ggplot2`.

- Loading [`tidyverse`]() also loads parts of `magrittr`, notably the _pipe_ `%>%`

- `r fontawesome::fa("hand-point-right")` `r fontawesome::fa("r-project")` `4.xx` offers a new pipe `|>`

]

???

Insist on the difference between _installing_ and _loading_ a package

- How do we get the list of installed packages?
- How do we get the list of loaded packages?
- Which objects are made available by a package?

`pacman::p_load()` kills two birds with one stone

---

### Have a look at `gapminder` dataset

.fl.w-30.pa2[

- `gapminder` is a table, like a spreadsheet table, or a relational database table

- In `r fontawesome::fa("r-project")` parlance, `gapminder` is a `data.frame`

- It is also a `tibble` in modern `tidyverse` parlance

- `r fontawesome::fa("bullhorn")` `data.frame`-like structures show up in every corner of Data Science

]

.fl.w-70.pa2[

```{r bare_gapminder}
gapminder %>% 
  glimpse(width=50)
```

]

---

### What's in a data frame/tibble?

A table has a _schema_: a list of named _columns_, each  with a given type

A table has a _content_: _rows_. Each row is a collection of items, corresponding
to the columns

`glimpse()` allows to see the schema and the first rows

`head()` allows to see the first rows

---

### Get a feeling of the dataset

.fl.w-30.pa2.f6[

```{r stratified_sampling}
tmp <- gapminder %>% 
  group_by(continent) %>%  #<<
  slice_sample(n=2)        #<<
```

]

.fl.w-70.pa2[

```{r stratified_sampling_bis, echo=FALSE}
tmp
```

]

---
exclude: true 

### A tidy table


???

What's make a table tidy?

- [ ] No/Few redundancies (what does it mean)
  - [ ] Is the `gapminder` table redundant?
  - [ ] ....


---
exclude: true

### Gapminder tibble  (extract)

```{r gapminder_td, eval=FALSE}
tt <- gapminder %>%
  filter(year==2002)  %>%
  mutate(pop = stringr::str_c(round(pop/1e6,1), 'M')) %>%
  DT::datatable(options=list(pageLength=8,
                dom = 't'),
                rownames = FALSE) %>%  #<<
  DT::formatCurrency('gdpPercap', digits=0) %>%      #<<
  DT::formatRound('lifeExp', 1) #<<
```



---
exclude: true


```{r gapminder_td_bis, echo=FALSE}

```

???

- [ ] Formatting instruction should not be wired in the table fed to `DT`
- [ ] reorder columns
- [ ] order countries by decreasing population size
- [ ] right-align `pop` column
- [ ] make provision for caption

---
###  `r fontawesome::fa("database")` $\sigma$:  Picking one year of data

.fl.w-30.pa2[

```{r gapminder_filtered, echo=TRUE}
gapminder_2002 <-
  gapminder %>%
  filter(year==2002) #<<
```

]

.fl.w-70.pa2.f6[

```{r gapminder_filtered_glimpse, echo=FALSE}
gapminder_2002
```

]


---

In the parlance of Relational Algebra, `filter` performs  a _selection_ of rows

[dplyr docs](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/filter){target="_blank"}

> find rows/cases where conditions are true

---
exclude: true

```{r flip_filter, include=FALSE}
gapminder %>%
  filter(year==2002) %>%  #ROTATE
  filter(continent == "Europe") %>%  #ROTATE
  filter(year==2002, continent == "Europe") %>%  #ROTATE
  filter(year %in% seq(2002,2008), continent %in% c("Africa", "Europe")) #ROTATE
```


```{r, echo=FALSE, eval=FALSE}
flipbookr:::embed_flipbook(chunk_name = "flip_filter",
               use_share_again = T,
               font_size = 120)
```

---
template: inter-slide

## Static plotting

---

### First attempt


.fl.w-40.pa2[

- Define a plot with respect to `gapminder_2002`

- Map variables `gdpPercap` and `lifeExp` to axes `x` and `y`

- For each row, draw a point at coordinates defined by the mapping

```{r bad, eval=FALSE, echo=TRUE}
p <- gapminder_2002 %>%
  ggplot() +    #<<
  aes(x=gdpPercap) + #<<
  aes(y=lifeExp) +  #<<
  geom_point() #<<

p
```

]


--

.fl.w-60.pa2[

```{r bad-out, ref.label="bad", echo=FALSE, fig.alt=""}

```

]



---

Recall that we are building a graphical object (a `ggplot` object) around a data frame. We supply _aesthetic mappings_ that can be either global
or bound to some _geometries_ or _statistics_ 

The global aesthetic mapping defines which columns are 

- mapped to which axes, 

- possibly mapped to colours, linetypes, shapes ...

Geometries and Statistics describe the building blocks of graphics

---

### What's missing here?

when comparing to the Gapminder demonstration, we can spot  that

-  [ ] colors are missing 

-  [ ] bubble sizes are all the same

-  [ ] titles and legends are missing

We will add layers to the graphical object to complete the plot

---

### Display more information


.fl.w-30.pa2[

- Map `continent` to color

- Map `pop` to bubble size

- Make point transparent by tuning `alpha` (avoid overplotting)

```{r better, echo=TRUE, eval=FALSE}
p <- p +
  aes(size = pop, #<<
      color = continent #<<
  ) +
  geom_point(alpha=.75)  #<<

p
```

]


.fl.w-70.pa2[

```{r better-out, ref.label="better", echo=FALSE}

```


]


---

### Scaling

.fl.w-30.pa2[

In order to pay tribute to Hans Rosling, we need to take care of two _scaling_ issues:

- the gdp per capita axis should be _logarithmic_

- the _area_ of the point should be proportional to the population

```{r scaling, echo=TRUE, eval=FALSE}
zoom_continent <- 'Europe'

p <- p +
  scale_x_log10(limits=c(100,5e4)) + #<<
  scale_size_area()  #<<

p
# p + ggforce::facet_zoom(xy= continent==zoom_continent,
#                         zoom.data = continent==zoom_continent)
```
]

.fl.w-70.pa2[


```{r scaling-out, ref.label="scaling", echo=FALSE}

```

]

---

### `r fontawesome::fa("brain")`

- Why is it important to use logarithmic scaling for gdp per capita?

- When is it important to use logarithmic scaling on some axis (in other contexts)?

- Why is it important to specify `scale_size_area()` ?

---

### In perspective


.fl.w-30.pa2[

- Add a plot title

- Make axes titles
  + explicit
  + readable

```{r title, echo=TRUE, eval=FALSE}
p <- p +
  labs(title=glue("Gapminder {min(p$data$year)}-{max(p$data$year)}"),
       x = "Yearly income per capita (US$)",
       y = "Life expectancy",
       caption= "From sick  and poor (bottom left) to healthy and rich (top right)")

```

]


.fl.w-70.pa2[

```{r title-out, ref.label="title", echo=FALSE}

```

]

---

### `r fontawesome::fa("brain")`

What should be the purposes of

- Title

- Subtitle

- Caption

---

### Theming using `ggthemes`


.fl.w-40.pa2[

- Theming

```{r theme_economist, echo=TRUE, eval=FALSE}
require("ggthemes")

p +
  theme_wsj()
```

A theme defines the _look and feel_ of plots

Within a single document, we should use only one theme

See [Getting the theme](https://ggplot2.tidyverse.org/reference/theme_get.html) for a gallery of available themes


]


.fl.w-60.pa2[

```{r theme_economist-out, ref.label="theme_economist", echo=FALSE}

```

]

???


- 
-


---



### Tuning scales


.fl.w-60.pa2[

```{r theme_scale, echo=TRUE, eval=FALSE}
neat_color_scale <-
      c("Africa" = "#01d4e5",
        "Americas" = "#7dea01" ,
        "Asia" = "#fc5173",
        "Europe" = "#fde803",
        "Oceania" = "#536227")
p <- p +
  scale_size_area(max_size = 15) + #<<
  scale_color_manual(values = neat_color_scale) #<<

p
```

Choosing a color scale is a difficult task

`viridis` is often a good pick

]



.fl.w-40.pa2[

```{r theme_scale-out, ref.label="theme_scale", warning=FALSE, echo=FALSE, message=FALSE}

```
]

---

### Zooming on a continent

.fl.w-40.pa2[

```{r zoom_africa, eval=FALSE, echo=TRUE}
require(ggforce) #<<

zoom_in <- 'Africa'

p_africa <- p +
  facet_zoom( #<<
    xy= continent==zoom_in, #<<
    zoom.data= continent==zoom_in #<<
    ) #<<
```

]


.fl.w-60.pa2[

```{r zoom_africa-out, ref.label="zoom_africa", warning=FALSE, echo=FALSE, message=FALSE}

```

]

???

Look for journal themes/ WSJ themes

Wes Anderson themes


---

```{r theme_scale-out-bis, ref.label="theme_scale", warning=FALSE, echo=FALSE, message=FALSE, out.width="90%"}
```




---
name: recap 
template: inter-slide


## Recap 


???


---

### Building a graphical object layer by layer

```{r plot_gap_minder_2002, eval=FALSE, fig.cap="Gapminder 2002 layer by layer"}
gapminder %>% 
  filter(year == 2002) %>%
   ggplot() +
   aes(x = gdpPercap) +
   aes(y = lifeExp) +
   aes(size = pop) +
   aes(color = continent) +
   geom_point(alpha=.5) +
   scale_x_log10() +
   scale_size_area(max_size = 15,
                  labels= scales::label_number(scale=1/1e6,
                                               suffix=" M")) +
   scale_color_manual(values = neat_color_scale) +
   theme_minimal() +
   theme(legend.position = "none") +
    labs(title= glue("Gapminder  {min(gapminder$year)}-{max(gapminder$year)}"),
         x = "Yearly Income per Capita",
         y = "Life Expectancy",
       caption="From sick  and poor (bottom left) to healthy and rich (top right)")

```

???


---

### In action 

`r chunk_reveal("plot_gap_minder_2002")`


???



---
template: inter-slide

## Facetting

---

So far we have only presented one year of data (2002)

Rossling used an animation to display the flow of time

If we have to deliver a printable report, we cannot rely on animation, 
but we can rely on _facetting_ 

Facets are collections of small plots constructed in the same way 
on subsets of the data

We add a layer to the graphical object using `facet_wrap()` 

As all rows in `gapminder_2002` are all related to `year` 2002


---

```{r facet, eval=FALSE, echo=TRUE}
p <- p +
  scale_x_log10(limits=c(100, 50000)) +
  scale_size_area(max_size = 15,
                  labels= scales::label_number(scale=1/1e6, suffix=" M")) +
  scale_color_manual(values = neat_color_scale) +
  guides(color = guide_legend(title = "Continent",
                              override.aes = list(size = 5),
                              order = 1),
         size = guide_legend(title = "Population",
                             order = 2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  facet_wrap(vars(year), ncol=6)  #<<

p
```


---


```{r facet-out, ref.label="facet", out.width="80%", echo=FALSE, message=FALSE, warning=FALSE}

```


---

We need to rebuild the graphical object along the same lines (using the same graphical pipeline) but starting from the whole `gapminder` dataset

Should we do this using _cut and paste_?

---

### Don't Repeat Yoursel (DRY)

> Abide to the DRY principle using operator `%+%`: the `ggplot2` object `p` can be fed with another dataframe and all you need is proper facetting.

```{r dryit, message=FALSE, warning=FALSE}
p  <- (p %+% gapminder)

p <- p + 
  labs(title=glue("Gapminder {min(p$data$year)}-{max(p$data$year)}"))

p
```


???



---

```{r dryit-out, ref.label="dryit", out.width="80%", echo=FALSE, message=FALSE, warning=FALSE}

```

---
template: inter-slide

## Animate for free with `plotly`

---

### From `r fontawesome::fa("r-project")` to `D3.js`

```{r animate, warning=FALSE, eval=FALSE, message=FALSE, out.width="80%"}
q <- filter(gapminder, FALSE) %>%
   ggplot() +
   aes(x = gdpPercap) +
   aes(y = lifeExp) +
   aes(size = pop) +
   aes(text = country) +
   aes(fill = continent) +
   aes(frame = year) +   #<<
  geom_point(alpha=.5, colour='black') +
  scale_x_log10() +
  scale_size_area(max_size = 15,
                  labels= scales::label_number(scale=1/1e6,
                                               suffix=" M")) +
  scale_fill_manual(values = neat_color_scale) +
  theme(legend.position = "none") +
  labs(title= glue("Gapminder  {min(gapminder$year)}-{max(gapminder$year)}"),
       x = "Yearly Income per Capita",
       y = "Life Expectancy",
       caption="From sick  and poor (bottom left) to healthy and rich (top right)")


(q %+% gapminder) %>%
  plotly::ggplotly(height = 500, width=750)     #<<
```

---

```{r}
(p +
 facet_null() +
 aes(frame=year)) %>%
 plotly::ggplotly(height = 500, width=750)
```

???

---

```{r animate-out, ref.label="animate", echo=FALSE, message=FALSE, warning=FALSE}


```


---

### Zooming on Europe (tweaking code)

```{r animate-europe, warning=FALSE, eval=FALSE, message=FALSE, out.width="80%"}
target_continent <- 'Europe'
(
  q %+%
  filter(gapminder, continent==target_continent) +
  ggtitle(glue("Gapminder on {target_continent}"))
) %>%
  plotly::ggplotly(height = 500, width=750)
```

---

### Zooming on Europe

```{r animate-europe-out, ref.label="animate-europe", echo=FALSE, message=FALSE, warning=FALSE}


```


---
template: inter-slide


```{r include_plotly_com}
knitr::include_url("https://plotly.com/r/")
```

---
template: inter-slide



```{r include_course}
knitr::include_url("https://www.youtube.com/embed/3n9nASHg9gc")
```

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

---
template: inter-slide


```{r include_course_2}
knitr::include_url("https://www.youtube.com/embed/RPFh3y9UAX4")
```


---

## Animate with `gganimate`

---

```{r  gap_gganimate, warning=FALSE, message=FALSE}
t_range <- with(gapminder, c(min(year), max(year)))
p +
  facet_null() + #<<
  gganimate::transition_time(year,  #<<
                             range = t_range) +#<<
  labs(subtitle="Year {frame_time}")

```

---
exclude: true 

### Permanence



---
exclude: true 

## Interactivity

---
exclude: true 

- `ggplotly`

- `highcharter`

???

[raukr](https://nbisweden.github.io/raukrtemplate/presentation_demo.html#1)

---
exclude: true 

### Under the hood

`htmltools`

`htmlwidgets`


???

- `facet_zoom()` on a continent (from `ggfortify`)

- Animation is not interaction!

- `shiny/flex_dashboard`

-

---

```{r child='closing_slide.Rmd'}

```
