---
title: "Univariate statistics"
subtitle: "EDA Master I, MIDS & MFA"
author: "Stéphane Boucheron"
institute: "Université de Paris"
date: "2020/11/20 (updated: `r Sys.Date()`)"
params:
  curriculum: "Master I MIDS & EDA"
  coursetitle: "Analyse Exploratoire des Données"
  lecturer: "Stéphane Boucheron"
  homepage: "http://stephane-v-boucheron.fr/courses/isidata/"
  curriculum_homepage: "https://master.math.univ-paris-diderot.fr/annee/m1-isifar/"
output:
  xaringan::moon_reader:
    css: ["header-footer.css",  "xaringan-themer.css", "custom-callout.css"]
    lib_dir: libs
    seal: false
    includes:
      in_header:
        - 'toc.html'
    nature:
      nature:
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
name: inter-slide
class: left, middle, inverse

{{ content }}

---
name: layout-general
layout: true
class: left, middle

```{r setup, child="loaders_fixers.Rmd", echo=FALSE, message=FALSE, warning=FALSE}

```

---

```{r child='title_slide.Rmd'}

```

---
exclude: true
template: inter-slide

# Univariate Statistics

### `r Sys.Date()`

#### [`r params$curriculum`](`r params$homepage_curriculum`)

#### [`r params$coursetitle`](`r params$homepage`)

#### [`r params$lecturer`](http://stephane-v-boucheron.fr)


---
template: inter-slide

## `r fontawesome::fa("map", fill="white")`

### [Samples, observations an variables ](samples)

### [Summary statistics](numerical-summaries)

### [Robust summary statistics](robust)

### [Boxplots](boxplots)

### [Histograms](histograms)

### [Cumulative distribution function](ecdf)

### [Quantile function](quantile-function)



---
template: inter-slide
name: samples

## Samples, observations an variables 


---

### `r fontawesome::fa("plane-departure")` `flights`dataset from `nycflights` package

.fl.w-50.pa2[
```{r }
pacman::p_load("nycflights13")
```

> On-time data for all flights that departed NYC (i.e. JFK, LGA or EWR) in 2013.

`flights` is a multivariate _sample_ 

Each `row` corresponds to an _observation_

Each column corresponds to a _measurement_ 

]

.fl.w-50.pa2.f6[

```
Columns: 19
$ year           <int> 2013,…
$ month          <int> 1, 1,…
$ day            <int> 1, 1,…
$ dep_time       <int> 517, …
$ sched_dep_time <int> 515, …
$ dep_delay      <dbl> 2, 4,…
$ arr_time       <int> 830, …
$ sched_arr_time <int> 819, …
$ arr_delay      <dbl> 11, 2…
$ carrier        <chr> "UA",…
$ flight         <int> 1545,…
$ tailnum        <chr> "N142…
$ origin         <chr> "EWR"…
$ dest           <chr> "IAH"…
$ air_time       <dbl> 227, …
$ distance       <dbl> 1400,…
$ hour           <dbl> 5, 5,…
$ minute         <dbl> 15, 2…
$ time_hour      <dttm> 2013…
```

]




---
### Definition(s)

A _sample_ is a sequence of items of the same type

In `r fontawesome::fa("r-project")`, if the type matches a basetype, a sample is typically represented using a `vector`

A _numerical_ (quantitative) sample is just a sequence of numbers (either integers
or floats), it is represented by a `numeric` vector.

A _categorical_ (qualitative)  sample is a sequence of values  from  some predefined finite set

In `r fontawesome::fa("r-project")`, categorical samples are handled using a special class: `factor`

---

In this lesson, we are concerned with _univariate statistics_, that
is either with

- _numerical samples_,

- _categorical samples_ taken from some finite (or seldom countable) set


`r fontawesome::fa("exclamation-triangle")` The boundary between qualitative and quantitative is not always straightforward


---

### `flights` dataset

- Columns `carrier`, `dest`, `origin` are considered categorical

```{r}
levels(factor(flights$origin))
```



- Columns `arr_delay`, `dep_delay`, distance are numerical

- Column `time_hour`, ... contains timestamps 


---


We will learn about:

- numerical summaries

- graphical displays

for univariate samples


---
class: middle, center, inverse
name: numerical-summaries

## Summary statistics

---

### Numerical summaries for quantitative variables

- __Mean__:

$$\overline{X}_n = \frac{1}{n}\sum_{i=1}^n x_i \qquad  \textbf{(location)}$$

- **Standard deviation**:

$$s^2 = \frac{1}{n-1} \sum_{i=1}^n \big(x_i - \overline{X}_n\big)^2 \quad \text{or}\quad s^2 = \frac{1}{n} \sum_{i=1}^n \big(x_i - \overline{X}_n\big)^2\qquad\textbf{(dispersion)}$$


---

### Beyond location et dispersion

- **Skewness**

$$\frac{\frac{1}{n} \sum_{i=1}^n \big(x_i - \overline{X}_n\big)^3}{s^3}$$

- **Kurtosis**

$$\kappa = \frac{\frac{1}{n} \sum_{i=1}^n \big(x_i -\overline{X}_n\big)^4}{s^4} - 2$$

---

### Numerical summaries at work


```{r}
s <- rnorm(1000)   # a random sample from the standard normal distribution

mean(s) - sum(s)/length(s)   #<<

sd(s) - sqrt(sum((s-mean(s))^2)/(length(s)-1))   #<<

mean((s- mean(s))^4)/sd(s)^4 -3   #<<
```

---

### Summarizing `arr_delay`

.fl.w-40.pa2.f6[

### One at a time

```{r}
mean(flights$arr_delay)
mean(flights$arr_delay, na.rm=T) #<<
sd(flights$arr_delay)
sd(flights$arr_delay, na.rm=T) #<<      
```
]


.fl.w-60.pa2.f6[

### All inclusive

```{r, warning=FALSE}
summary(flights$arr_delay) %>%   #<<
  broom::tidy() %>% 
  knitr::kable(digits = 2,
  caption = "Summary statistics for nycflights13 arrival delay")

sum_funs <- list(mean, 
  sd, median, IQR)

purrr::map_dbl(sum_funs, #<<
  ~ .x( flights$arr_delay, na.rm=T)) #<<
```

`r fontawesome::fa("hand-point-right")` NAs matter!

]
---

### With `dplyr::summarise`

```{r}
flights %>% 
  select(- year, -month, -day, -hour, -minute , -time_hour, -ends_with("time"), -flight) %>% 
  summarise(across(where(is.numeric),   #<<
                   .fns=list("mean"=mean, "median"=median , "IQR"=IQR), #<<
                   na.rm=T, #<<
                   .names="{.col}_{.fn}")) %>%  #<<
  pivot_longer(cols = everything()) %>% 
  knitr::kable(digits=2)
```


---
exclude: true

### What if some data are missing?

```{r}
s <- rnorm(10)
s[c(2, 3, 7)] = NA_real_
s
```

```{r}
mean(s)   #<<
mean(s, na.rm = TRUE)  #<<
mean(s[-c(2, 3, 7)])
```



---

### Missing values and trimming

The code for `mean.default` goes beyond computing the weighted mean

.f6[

```{r, eval=FALSE}
function (x, trim = 0, na.rm = FALSE, ...)
{
    if (!is.numeric(x) && !is.complex(x) && !is.logical(x)) {
        warning("argument is not numeric or logical: returning NA")
        return(NA_real_)
    }
    if (na.rm)    #<<
        x <- x[!is.na(x)]    #<<
    if (!is.numeric(trim) || length(trim) != 1L)
        stop("'trim' must be numeric of length one")
    n <- length(x)
    if (trim > 0 && n) {
        if (is.complex(x))
            stop("trimmed means are not defined for complex data")
        if (anyNA(x))
            return(NA_real_)
        if (trim >= 0.5)
            return(stats::median(x, na.rm = FALSE))
        lo <- floor(n * trim) + 1      #<<
        hi <- n + 1 - lo               #<<
        x <- sort.int(x, partial = unique(c(lo, hi)))[lo:hi]  #<<
    }
    .Internal(mean(x))  #<< 
}

```
]


---

### `r fontawesome::fa("brain")`

- What happens when `na.rm` is `TRUE`?

- What are the possible values of argument `trim`?

- If `trim` belongs to $(0, 1/2)$ what happens?

- Which lines actually compute the (uniformly) weighted mean?

???


- The first `if (...) {...}` is an example of defensive programming

- `if (na.rm) { ... }` handles missing data when the default behavior is overriden.
`NA` are filtered out

- When positive, parameter  `trim` allows us to compute _trimmed means_

Depending on optional arguments, fucntion `mean` allows to compute either arithmetic means, or trimmed (arithmetic) means

---

### `r fontawesome::fa("hand-point-right")`

- In a _tidy_ sample, _missing values_ are explicitely pointed as missing information using special objects `NA`, `NA_integer_`, `NA_real_`, ...

- To compute summary statistics when facing missing values,  `na.rm` has to be `TRUE`.

- What is the class of `NA`, `NA_integer_`, `NA_real_`?


---

### `r fontawesome::fa("brain")`

Why is `var(sample)` defined by 

$$\sum_{i=1}^n \frac{1}{n-1} (x_i - \overline{X}_n)^2$$

when $\texttt{sample}=x_1, \ldots, x_n$?

---
template: inter-slide
name: robust

## Robust summary statistics

---
exclude: true
### `summary(...)`

```{r}
summary(s)
```

---

- **Order statistics**: $x_{1:n} \leq x_{2:n} \leq \ldots \leq x_{n:n}$

- **Median**  $x_{\lfloor n/2\rfloor+1:n}$ if $n$ is odd,
$(x_{n/2:n}+ x_{n/2+1:n})/2$ if $n$ is even

- **Quartiles** $x_{\lfloor n/4\rfloor:n}$ and $x_{\lfloor 3 n/4\rfloor:n}$

- **Quintiles** $x_{\lfloor k n/5\rfloor:n}$ pour $k \in 1, \ldots, 4$

- **Deciles** $x_{\lfloor k n/10\rfloor:n}$ pour $k \in 1, \ldots, 9$

- **Percentiles**

- ...


---

### `r fontawesome::fa("brain")`

Compute the derivative of 

- the mean

- the median 

- the trimmed mean 

with respect to $x_{(n)}$




---
exclude: true

Why _robustness_?

.bg-light-gray.b--light-gray.ba.bw2.br3.shadow-5.ph4.mt5[

'Est etincidunt non adipisci quaerat sed. Ipsum adipisci numquam neque ipsum modi neque modi. Amet velit tempora aliquam etincidunt. Quaerat neque labore velit dolorem dolore eius ipsum. Modi sed dolorem aliquam dolorem non neque. Dolor velit porro etincidunt magnam aliquam labore. Sed quisquam dolore tempora dolor ipsum sit est. Porro numquam adipisci aliquam. Amet labore etincidunt dolore porro.'

]

---
exclude: true


Let $x_1, \ldots,  x_n$ be a numerical sample.

```{r}
rapportools::kurtosis
```

Check that the kurtosis and the skewness are translation invariant.

---
template: inter-slide
name: boxplots

## Boxplots

---


.bg-light-gray.b--light-gray.ba.bw2.br3.shadow-5.ph4.mt5[

In descriptive statistics, a box plot or boxplot is a method for graphically depicting groups of numerical data through their _quantiles_ (more specifically _quartiles_)

Box plots also have lines extending from the boxes (__whiskers__) indicating variability outside the upper and lower quartiles, hence the terms _box-and-whisker plot_ and box-and-whisker diagram

_Outliers_ may be plotted as individual points

Box plots are _non-parametric_: they display variation in samples of a statistical population without making any assumptions of the underlying statistical distribution (though Tukey's boxplot assumes symmetry for the whiskers and normality for their length)

The spacings between the different parts of the box indicate the degree of _dispersion_ (spread) and _skewness_ in the data, and show _outliers_


.tr[
[Wikipedia](https://en.wikipedia.org/wiki/Box_plot)
]

]

---

> Tukey's boxplot assumes symmetry for the whiskers and normality for their length

What does this mean?

Assume that we _boxplot_ the $\mathcal{N}(\mu, \sigma^2) \sim \mu + \sigma \mathcal{N}(0, 1)$ distribution instead of an empirical distribution.

Then:

- The median is equal to $\mu$ because of the symmetry of $\mathcal{N}(0,1)$,

- The quartiles are $\mu - \sigma \Phi^{\leftarrow}(3/4)$ and $\mu + \sigma \Phi^{\leftarrow}(3/4)$.

- The IQR is $2 \sigma  \Phi^{\leftarrow}(3/4) \approx 1.35\times \sigma$


- The default length of the whiskers (under `geom_boxplot`) is
$1.5 \times \text{IQR}$

- For a Gaussian population, $1.5 \text{IQR} = 3 \sigma \Phi^{\leftarrow}(3/4)$

- In a Gaussian population, the probability that a sample point lies outside the whiskers is $2 \times (1 - \Phi(4 \times \Phi^{\leftarrow}(3/4))) \approx 7/1000$


---
class: top, left

### Boxplot of a Gaussian sample



```{r normal-boxplot-label, message=FALSE, warning=FALSE, out.width="70%", fig.show='hide'}
n <- 100
fake_data <- tibble(normal=rnorm(n),t=rt(n, df=3))
title <- stringr::str_c("Standard Gaussian sample, size : ", n)

fake_data %>%
  ggplot(mapping = aes(x=1L, y=normal)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 2, coef=1.5) + #<<
  ggtitle(title) +
  ylab("") +
  ggplot2::theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank()
  )

summary(fake_data$normal)
```


.plot-callout[
![gauss-boxplot-label](`r knitr::fig_chunk("normal-boxplot-label", "png")`)
]

---

### Boxplot of a Gaussian sample

.fl.w-60.pa2[

![gauss-boxplot-label](`r knitr::fig_chunk("normal-boxplot-label", "png")`)

]

???

Playing with boxplots

- Tune whiskers lengths using `coef`

- Jitter to spot the whole

---

If we _boxplot_ a Gaussian sample, on average, $4\%$ of the points lie outside the whiskers. Such points are called _extremes_

Deciding which points should be considered as extremes is partly a matter of taste

Boxplots provide a visual assessment of the departure of a sample from normality/Gaussianity

[`geom_boxplot` documentation](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)

???

Point out the distinction between outliers and extremes


---


### Boxplot of a non-Gaussian sample




The sampling distribution ( $t$-distribution with $3$ degrees of freedom) is _heavy-tailed_

Some sample points lie far away from the whiskers

```{r non-gauss-boxplot-label, fig.show='hide', out.width=.8}
title <- stringr::str_c("Student sample with 3 df, size : ", n)
ggplot(fake_data,
       aes(y=t, x=1L)) +
  geom_boxplot(outlier.shape=2, outlier.colour = 'red', outlier.alpha=.5, coef=1.5) + #<<
  ylab("") +
  ggtitle(title) +
  ggplot2::theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank()
  )
summary(fake_data$t)
```



.plot-callout[

![non-gauss-boxplot-label](`r knitr::fig_chunk("non-gauss-boxplot-label", "png")`)

]

---

### Boxplot of a non-Gaussian sample

.fl.w-60.pa2[

![non-gauss-boxplot-label](`r knitr::fig_chunk("non-gauss-boxplot-label", "png")`)

]

---

### Plotting two boxplots side by side

.fl.w-50.pa2[

```{r two-boxplots, echo=TRUE, eval=FALSE}
n <- 1000

tibble(normal=rnorm(n), 
       t=rt(n, df=3)) %>%
  pivot_longer(cols=c(normal, t),  #<<
               names_to='Distribution') %>%  #<<
  ggplot() + #<<
  aes(x=Distribution, y=value) +   #<<
  geom_boxplot(outlier.alpha = .5,   #<<
               outlier.colour = 'red',
               outlier.shape = 2) +
  ggtitle("Gaussian versus heavy-tailed samples")
```
]

.fl.w-50.pa2[

```{r two-boxplots-out, echo=FALSE,  ref.label="two-boxplots"}

```

]

---

### Boxplots for normal and heavy-tailed distributions

With usual whiskers length

- In Gaussian samples: 
  + very few sample points outside the whiskers (few extremes/outliers), 
  + extremes lie close to the whiskers endpoints

- In heavy-tailed distributions: 
  + many sample points outside the whiskers, and 
  + some extremes may lie far far away


---
exclude: true

### Boxplotting Titanic Fare prices

.fl.w-60.pa2[

```{r titanic-fares-boxplot-label, echo=FALSE}
title <- stringr::str_c("Titanic Fare prices")

p <- ggplot(tit,
       aes(y=Fare, x=1L)) +
  geom_boxplot(outlier.shape=2, outlier.colour = 'red', outlier.alpha=.5, coef=1.5) + #<<
  ylab("") +
  scale_y_continuous(
  labels = scales::number_format(scale = 1,
                                 suffix = " £"))+
  ggtitle(title) +
  ggplot2::theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank()
  )

p
# summary(tit$Fare)
```
]


.fl.w-40.pa2[

Titanic fare prices distribution exhibits a non-Gaussian behavior
`r fontawesome::fa("smile-wink")`

The median looks much closer to the first quartile than to the third quartile (asymmetry)

There are many extreme points

Extreme lie far away above the whiskers

]

---

### Boxplots for delays 

.fl.w-40.pa2.f6[

```{r boxplots-delays, echo=TRUE, eval=FALSE}
flights %>% 
  select(
    ends_with("_delay")
    ) %>% 
  pivot_longer(
    cols=ends_with("_delay"),
    names_pattern = "(.*)_delay", 
    values_to = "delay"
    ) %>% 
  filter(
    abs(delay) < 120
    ) %>% 
  ggplot() +
  aes(x=name, y=delay) +
  geom_boxplot()
```

]

.fl.w-60.pa2[

```{r boxplots-delays-out, ref.label="boxplots-delays", echo=FALSE, eval=TRUE}

```



]



---
exclude: true
template: inter-slide
name: violinplots

## Violinplots

---
exclude: true

.fl.w-60.pa2[

```{r titanic-fares-violinplot-label, echo=FALSE}
title <- stringr::str_c("Titanic Fare prices")

p <- ggplot(tit,
       aes(y=Fare, x=1L)) +
  geom_violin(aes(x=Pclass)) + #<<
  ylab("") +
  scale_y_continuous(
  labels = scales::number_format(scale = 1,
                                 suffix = " £"))+
  ggtitle(title) +
  xlab("Passenger class") +
  ggplot2::theme(
  #axis.text.x = element_blank(),
  # axis.ticks.x = element_blank()
  )

p
# summary(tit$Fare)
```
]


.fl.w-40.pa2[

Titanic fare prices distribution exhibits a non-Gaussian behavior
`r fontawesome::fa("smile-wink")`

They are a mix of three populations that also exhibit non-Gaussian behavior
]




---
exclude: true

.bg-light-gray.b--light-gray.ba.bw2.br3.shadow-5.ph4.mt5[

'Est etincidunt non adipisci quaerat sed. Ipsum adipisci numquam neque ipsum modi neque modi. Amet velit tempora aliquam etincidunt. Quaerat neque labore velit dolorem dolore eius ipsum. Modi sed dolorem aliquam dolorem non neque. Dolor velit porro etincidunt magnam aliquam labore. Sed quisquam dolore tempora dolor ipsum sit est. Porro numquam adipisci aliquam. Amet labore etincidunt dolore porro.'

]



---
template: inter-slide
name: histograms

## Histograms

---

### Definition(s)

.fl.w-40.pa2[

`r fontawesome::fa("hand-point-right")` A histogram defines a piecewise constant probability density function

A histogram is defined by the number and position of the breaks (bins)


A histogram may be regular or not 

]

.fl.w-60.pa2[

```{r histogram-gauss, echo=FALSE}
tibble(x=rnorm(1000, mean = 2)) %>% 
  ggplot() +
  aes(x=x) +
  geom_rug(alpha=.1) +
  geom_histogram(aes(y=..density..), 
                 alpha=.5, 
                 fill="white", 
                 colour="black", 
                 bins=30) +
  stat_function(linetype="dashed", 
                fun=dnorm, 
                args=list(mean=2, sd=1)) +
  annotate("text", x=4.5, y=.3, 
           label="Gaussian density with mean 2") +
  ggtitle("Histogram for Gaussian sample") +
  labs(caption="Sample size=1000, 30 bins")
```



]


---



`r chunk_reveal("histogram-gauss")`

---

### Construction

- Binning the sample values
  + Picking the right (?) number of bins
  + Placing the breaks
  
- 


---

### Histograms for (trimmed) delay distributions

.fl.w-40.pa2.f6[
```{r  histograms-arr_delay, eval=FALSE}
flights %>% 
  select(
    ends_with("_delay")
    ) %>% 
  pivot_longer(
    cols=ends_with("_delay"),
    names_pattern = "(.*)_delay", 
    values_to = "delay"
    ) %>% 
  filter(
    abs(delay) < 120
    ) %>% 
  ggplot() +
  aes(x=delay, #<<
      y=..density..,  #<<
      fill=name) +   #<<
  geom_histogram(alpha=.3,  #<< 
                 bins=30)  #<<
```
]

.fl.w-60.pa2.f6[

```{r histograms-arr_delay-out, echo=FALSE, ref.label="histograms-arr_delay"}

```

]

---
exclude: true

.bg-light-gray.b--light-gray.ba.bw2.br3.shadow-5.ph4.mt5[

'Est etincidunt non adipisci quaerat sed. Ipsum adipisci numquam neque ipsum modi neque modi. Amet velit tempora aliquam etincidunt. Quaerat neque labore velit dolorem dolore eius ipsum. Modi sed dolorem aliquam dolorem non neque. Dolor velit porro etincidunt magnam aliquam labore. Sed quisquam dolore tempora dolor ipsum sit est. Porro numquam adipisci aliquam. Amet labore etincidunt dolore porro.'

]

---
template: inter-slide
name: ecdf

## Cumulative distribution function

---


.bg-light-gray.b--light-gray.ba.bw2.br3.shadow-5.ph4.mt5[

$P$ be a probability distribution over $\mathbb{R}$ (endowed with the
Borelian $\sigma$-algebra)

The cumulative distribution function (CDF) $F$ is a function from $\mathbb{R}$ to $[0,1]$

$$F(x) = P(-\infty, x)$$

]


---

A real-valued sample defines a probability distribution, called the _empirical distribution_.
Each sample point is assigned a probability equal to its relative frequency.


The cumulative distribution function associated with this empirical distribution
is called the _empirical cumulative distribution function_ (ECDF).

---

Let $x_1, \ldots, x_n$ denote the points from a real sample (they may not be pairwise
distincts), let $P_n$ be the empirical distribution and $F_n$ be the ECDF.

$$\begin{array}{rl}  P_n(A) & = \frac{1}{n} \sum_{i=1}^n \mathbb{I}_{x_i \in A}  \\ F_n(z) & = \frac{1}{n} \sum_{i=1}^n \mathbb{I}_{x_i \leq z} \end{array}$$

---

Let $x_{1:n} \leq x_{2:n} \leq \ldots \leq x_{n:n}$ denote the _order statistics_
of the sample (that is the non-decreasing rearrangement of sample, with possible
repetitions).


The ECDF is easily computed from the order statistics:


$$F_n(z) = \begin{cases} 0           & \text{if } z < x_{1:n}               \\ \frac{k}{n} & \text{if } x_{k:n} \leq z < x_{k+1:n}\\ 1           & \text{if } x_{n:n} \leq z \end{cases}$$

---

Recall that ECDFs are determined  by order statistics (and conversely)


.bg-light-gray.b--light-gray.ba.bw2.br3.shadow-5.ph4.mt5[

### Proposition

The empirical cumulative distribution function $F_n$ is

- right-continuous over $\mathbb{R}$

- non-decreasing

- takes values in $\{ k/n , k \in \{0, 1, \ldots, n\}\}$

- is constant between consecutive order statistics

]

---

### Example

Any numerical dataset may be used to illustrate ECDFs.


For example,

`faithful`: a data frame with `272` observations on `2` variables concerning waiting time between eruptions and the duration of the eruption for the _Old Faithful_ geyser in Yellowstone National Park, Wyoming, USA.

- `eruptions`:	`numeric`,	Eruption time in minutess

- `wating`:	`numeric`	Waiting time to next eruption (in minutes)

> A closer look at `faithful$eruptions` reveals that these are heavily rounded times originally in seconds, where multiples of 5 are more frequent than expected under non-human measurement.


---

###  A glimpse at the faithfull data 

```{r, echo=TRUE}
data(faithful)
faithful %>%
  glimpse()
```



---



### Plotting an empirical CDF

.fl.w-third.pa2[

Eruptions durations

Under the hood, `ggplot` combines `geom_step`  and `ecdf` statistic


```{r  geyser-label, echo=TRUE, fig.show='hide'}
faithful %>%
  ggplot() +
  aes(x=eruptions) +
  geom_step(stat="ecdf") +   #<<
  ylab("ECDF") +
  xlab("Eruptions duration in minutes") +
  ggtitle("Old Faithful Geyser Data")
```

]

.fl.w-two-thirds.pa2[

![](`r knitr::fig_chunk('geyser-label', 'png')`)

]

---
exclude: false

### Computing the ECDF statistic


.fl.w-third.pa2[

ECDF plots may be used to compare two samples or to compare an empirical
cumulative distribution function with the cumulative distribution function
of some well-known probability distribution.

```{r  geyser-label-2, echo=TRUE, fig.show='hide'}
Fn <- ecdf(faithful$eruptions) #<<

x_grid <- seq(min(faithful$eruptions)-1,
              max(faithful$eruptions)+1,
              0.01)

df <- data_frame(x=x_grid,
                 Fn=Fn(x_grid))

df %>%
  ggplot() +
  aes(x=x, y=Fn) +
  geom_step(direction="hv")
```

]


.fl.w-two-thirds.pa2[

![](`r knitr::fig_chunk('geyser-label-2', 'png')`)

]


---
template: inter-slide
name: quantile-function

## Quantile function

---
exclude: false


.bg-light-gray.b--light-gray.ba.bw2.br3.shadow-5.ph4.mt5[

The _quantile function_ of a probability distribution over $\mathbb{R}$ is
the generalized, left-continuous inverse function of the cumulative
distribution function.

For $p \in (0,1)$

$$F_n^{\leftarrow}(p) = \inf \Big\{ z : F_n(z) \geq p\Big\}$$

$$F_n^{\leftarrow}(p) =  X_{k+1:n} \qquad \text{if } \qquad\frac{k}{n} < p \leq \frac{k+1}{n}$$

]


---


.fl.w-third.pa2[

In order to plot the quantile function, it is almost enough to exchange the
role of the axes in the ECDF plot.

```{r  eruptions-label, echo=TRUE, fig.show='hide'}
faithful[1:10,] %>%
  ggplot() +
  aes(x=eruptions) +
  geom_step(stat="ecdf", 
            direction='vh') +
  coord_flip() +   #<< 
  xlab("Quantiles") +
  ylab("Probabilities") +
  ggtitle("Old Faithful Geyser Data")
```

Why is it safe to set the `direction` argument in the call to `geom_step()`

]


.fl.w-two-thirds.pa2[

![](`r knitr::fig_chunk("eruptions-label", "png")`)


]


---
exclude: false

### Comparing sample quantiles with distributional quantiles

`r fontawesome::fa("bullhorn")`


When it comes to comparing the empirical distributions defined by two samples,
_quantile-quantile plots_ ( `qqplot` ) are favored.


- $F_n$ and $G_m$ : two ECDFs

- the quantile-quantile function is defined by

$$G_m^\leftarrow \circ F_n(z) = Y_{k+1:m}\qquad \text{if} \qquad \frac{k}{m}< F_n(z) \leq \frac{k+1}{m}$$


---


If $G_m = F_n$ (if the two samples are equal up to a permutation), then

$$F_n^\leftarrow \circ F_n(z) = X_{k+1:n} \text{if} \qquad X_{k:n} < z \leq X_{k+1:n}$$

The quantile-quantile plot lies above line $y=x$ and meets this line at every
unique order statistics.

`r fontawesome::fa("grin")` The departure of the quantile-quantile function from identity reflects
the  differences between   the two empirical distribution functions.

We can for example compare the first and second half of the `faithfull`
dataset.

---

### Defining an empirical quantile function

.pull-left[

Function `equantile()` returns a function

The returned function is _parametrized_ by argument
`sample`

.small[

- A function captures (encloses) the _environment_ in which it is created

- Function `equantile` creates a new _execution environment_ every time it is run. This environment is usually ephemeral, but here it becomes the _enclosing environment_ of the manufactured function

- The _enclosing environment_ of the manufactured function is an _execution environment_ of the function factory

]


]

.pull-right[

#### A quantile function factory

```{r}
equantile <- function(sample){
  sample <- sort(sample)
  m <- length(sample)

  function(p){     #<<
    ks <- pmax(1, ceiling(m * p)) #<<
    sample[ks]  #<<
  }
}


```

See [Advanced R programming](https://adv-r.hadley.nz/function-factories.html)
]

]


---

### QQ plot

.fl.w-40.pa2.f6[

```{r qq_delays, eval=FALSE}
not_cancelled <- flights %>% 
  filter(!is.na(arr_delay)) %>% 
  sample_n(100) %>% 
  arrange(dep_delay) %>% 
  filter(abs(arr_delay) < 120) 
  

nr <- nrow(not_cancelled)


not_cancelled %>% 
  ggplot() +
  aes(x=dep_delay, 
      y=quantile(arr_delay, 
                 seq(1, nr)/nr)
      ) +
  geom_point(size=.1) +
  coord_fixed() +
  geom_abline(slope=1, 
              intercept = 0, 
              alpha=.5) +
  ggtitle("QQ plot clipped arr_delay versus dep_delay") +
  ylab("arr_delay") +
  xlab("dep-delay")
```
]


.fl.w-60.pa2[

```{r qq_delays-out, echo=FALSE, ref.label="qq_delays"}

```


]


---
exclude: true

We can repeat this experiment after sample shuffling.

.fl.w-third.pa2[

```{r shuffle-faithful-label, echo=TRUE, fig.show='hide'}
unfaithful <- faithful[sample(seq(1,nrow(faithful)),
                               size=nrow(faithful),
                               replace = FALSE),
                        ]

first <- unfaithful[1:135,]
second <- unfaithful[-(1:135),]

Fn <- ecdf(first$eruptions)

Qm <- equantile(second$eruptions)

df <- data.frame(x= seq(min(first$eruptions)-1,
                        max(first$eruptions)+1, 0.01)) %>%
  dplyr::mutate(y = Qm(Fn(x)))

df %>%
  ggplot(mapping = aes(x=x, y=y)) +
  geom_step(direction='vh') +
  geom_abline(slope=1,
    intercept = 0,
    linetype='dashed')

```
]

.fl.w-two-thirds.pa2[

![](`r knitr::fig_chunk('faithful-shuffle-label','png')`)

]

---
exclude: true 

```{r, faithful-split-label-bis, echo=TRUE, fig.show='hide'}
first <- faithful[1:135,]
second <- faithful[-(1:135),]

Fn <- ecdf(first$eruptions)
Qm <- equantile(second$eruptions)

df <- data.frame(x= seq(min(first$eruptions)-1, max(first$eruptions)+1, 0.01)) %>%
  dplyr::mutate(y = Qm(Fn(x)))


df %>%
  ggplot() +
  aes(x=x, y=y) +
  geom_step(direction='vh') +
  geom_abline(slope=1,
              intercept = 0,
              linetype='dashed')

```


---

```{r child='closing_slide.Rmd'}

```
