
---

In this lesson, we further explore the connection between two qualitative variables, that is,
we attempt to go beyond Lesson [Bivariate statistics](#biv).

Our starting point is a qualitative
bivariate sample.

Consider the  celebrated `UCBAdmissions` dataset

According to `R` documentation,
this dataset is made of

> Aggregate data on applicants to graduate school at Berkeley for the six largest departments in 1973 classified by admission and sex.

This is a compilation of `r sum(datasets::UCBAdmissions)` application files.

For each application file,
three variables have been reported: the `department`, the `gender` of the applicant and whether
the applicant has been `admitted`

The dataset is a trivariate sample. In the sequel we make it
a bivariate sample by focusing on `Gender`  and `Admit`.

???

The dataset was gathered because
the admission process at UCB graduate schools was suspected to be biased against women


---

### Plotting two-way contingency tables

As in Lesson [Bivariate statistics](#biv), we start by collecting the 2-way _contingency table_ and
displaying a _mosaicplot_.

```{r, echo=FALSE}
UCBAdmissions <- datasets::UCBAdmissions

apply(UCBAdmissions, MARGIN = c(1,2), sum) %>%
  as_tibble(rownames = "Admit") -> ucb2
```

.panelset[

.panel[.panel-name[`mosaicplot`]

.fl.w-50.pa2[

```{r mosaic1, fig.height=4, fig.show='hide'}
mosaicplot(~ Gender + Admit,   #<<
           UCBAdmissions)      #<<
```

]

.fl.w-50.pa2[

![](`r knitr::fig_chunk("mosaic1", "png")`)

]

]

.panel[.panel-name[`mosaic` (from `vcd`)]


.fl.w-50.pa2[

```{r mosaic2, fig.height=4, fig.show='hide'}
apply(UCBAdmissions,
      MARGIN = c(1,2),
      sum) %>%
  as.table() %>%
  vcd::mosaic(formula=Admit~Gender,  #<<
              data=.)                #<<
```
]

.fl.w-50.pa2[

![](`r knitr::fig_chunk("mosaic2", "png")`)

]

]

.panel[.panel-name[`ggmosaic`]

.fl.w-50.pa2[

.f6[
If, rather than the contingency table,  we have the original data (a dataframe
with two categorical columns), we can visualize the contengincy table using
`geom_mosaic` from `ggmosaic`. Conforming to the `ggplot2` API, `geom_mosaic`
works on a native dataframe and compute _statistics_ from the dataframe, the
statistics are then used to draw the picture

```{r prep-mosaic3, echo=FALSE, eval=TRUE}
titanic_train <- readr::read_csv("DATA/TITANIC/train.csv",
                col_types = cols(Embarked = col_factor(levels = c("Q", "C", "S")),
                                 Pclass = col_factor(levels = c("1", "2", "3")),
                                 Sex = col_factor(levels = c("male", "female")),
                                 Survived = col_factor(levels = c("0", "1"))))
```

```{r mosaic3, fig.height=4, fig.show='hide'}
titanic_train %>%
  dplyr::select(Pclass, Embarked) %>%
  drop_na() %>%
  ggplot() +
  geom_mosaic(aes(x = product(Embarked, Pclass), fill=Embarked)) +
  labs(x= "Passenger class", y="Embarked") +
  scale_fill_viridis_d() +
  ggtitle("Titanic mosaic with tidyverse flavor")
```
]
]

.fl.w-50.pa2[

![](`r knitr::fig_chunk("mosaic3", "png")`)

]

]
]

???


This mosaicplot suggests that Admission and Gender are associated.

[ggmosaic](https://cran.r-project.org/web/packages/ggmosaic/vignettes/ggmosaic.html)

