---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

## Context: EpiCov investigation 

---


```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width="50%", fig.show='hold')
require(tidyverse)
require(vcd)
# require(plotly)
require(viridis)
require(FactoMineR)
require(factoextra)
require(FactoInvestigate)
#library(lobstr)
pacman::p_load("factoextra")
```

### Source

This notebook is based on

Reference: [Social inequalities in hostility toward vaccination against Covid-19](https://doi.org/10.1101/2021.06.07.21258461)

by Alexis Spire, Nathalie Bajos, Léna Silberzan , for the EPICOV study group



> In recent decades, France has appeared as a country particularly hostile to vaccination in general. When asked in November 2020 about the intention to take the Covid-19 vaccine, the French public showed, once again, reluctance. Therefore, France appeared as an ideal case study to analyze whether the hostility toward the Covid-19 vaccine has its own reasons or whether it is related to the reluctance to the principle of vaccination itself. Our objective was to determine the specificity of the social determinants of the intention to get the Covid-19 vaccine. Thanks to the use of a large random sample of the general population in France (86,000 individuals), the reluctant to Covid-19 vaccine could be clearly distinguished from the hesitant and the convinced, and thereby thoroughly analyzed. Our analysis highlighted a gendered reluctance toward vaccination in general but even more so regarding vaccination against Covid-19. It might refer to women being more concerned about the possible effects of an injection in their body, especially at the age of maternity and a differentiated socialization making them more sensitive than men to long-term risks and more apprehensive toward rapid technological change. We also found that people at the bottom of the social hierarchy, in terms of level of education, financial resources, and immigration status, were more likely to refuse the Covid-19 vaccine. Nevertheless, this reluctance was less prominent than for vaccination in general, reflecting the actual spread of the epidemic in various social milieux. Finally, our analysis showed that trust in the government’s actions was significantly associated with reluctance toward the Covid-19 vaccine, even more than toward vaccination in general.




---

### The EPICOV investigation

> The EpiCoV (Epidémiologie et Conditions de Vie) cohort was set-up in April 2020, with the general aim of understanding the main epidemiological, social and behavioural issues related to the Covid-19 epidemic in France. The survey was approved by the CNIL (French independent administrative authority responsible for data protection) on April 25th 2020 (ref: MLD/MFI/AR205138) and by the “Comité de protection des personnes” (French equivalent of the Research Ethics Committee) on April 24th. The survey also obtained an agreement from the _Comité du Label de la statistique publique_, proving its adequacy to statistical quality standards.

See [https://doi.org/10.1101/2021.02.24.21252316](https://doi.org/10.1101/2021.02.24.21252316) for more the EpiCov _cohort_


> This study was based on a large-scale random survey of 107,808 people conducted between October 26 and December 9, 2020, a pivotal time, as Pfizer announced on November 9, 2020, that it would be able to produce a $90\%$ effective vaccine on a large scale.



---

### Demographic variables 


> To describe the sample, six sociodemographic variables were considered: _age_, _gender_, _ethno-racial status_ (based on _migration history_), social class (based on _current or last occupation_), standard of living (based on _decile of household income_ per consumption unit), and _formal education level_. Ethno-racial status was defined by combining the criteria of place of birth, nationality, and status of the individual and both parents.

> The analysis was conducted from an intersectional perspective [10] that simultaneously took into account gender, class, age, and ethno-racial social characteristics, as well as respondents' level of trust in the government.


---

### Attitudinal variables: general versus specific vaccination hesitancy

> To study attitudes toward vaccination in the EpiCoV survey in November 2020, two questions were available.

> - About vaccination in general: Are you strongly; somewhat; somewhat not; or not at all in favor of vaccinations in general?

> - About the Covid-19 vaccine:  If a free vaccine against coronavirus was offered by the Sécurité Sociale, would you be willing to get vaccinated: Yes probably; yes maybe; probably not; certainly not; or you do not know.

- [ ] Attitudes towards vaccination are reported on a Likert/rating scale. Is there any difference between the rating scales for the two questions?
- [ ] If you find any difference, can you guess the motivation?


### Rating scales 

[Likert scale/Rating scale](https://en.wikipedia.org/wiki/Likert_scale)



```{r, echo=FALSE}
lkt_will <- c("Yes Probably", "Yes Maybe", "Do not know", "Probably Not", "Certainly Not")
lkt_general <- c("Strongly in Favor", "Somewhat in Favor", "Somewhat Not in Favor", "Not at all in Favor")

xtab_general_covid <- matrix(
  c(16062, 3411, 993, 468, 386,
    12607, 16190, 6234, 3901, 2705,
    928, 3321, 3408, 3947, 3242,
    227, 524, 1344, 1144, 4723),
  nrow = 4, ncol = 5,
  byrow = TRUE,
  dimnames = list("vaccine"=lkt_general, "covid vaccine"=lkt_will)) %>%
  as.table()
```

---

### Mosaiplot vaccination against covid vaccination

```{r, echo=FALSE}
mosaicplot(xtab_general_covid, shade = TRUE)
```


---

### Mosaic plot (bis)


```{r, echo=FALSE}
mosaic(xtab_general_covid, shade = TRUE)
```

---
template: inters_slide

## Correspondence analysis

---

### Screeplot

```{r, echo=FALSE}
# out.width="50%", fig.show='hold'
ca_xtab_general_covid <- CA(xtab_general_covid,graph = FALSE)

#fviz_eig(ca_xtab_general_covid)

ca_xtab_general_covid$eig %>%
  as_tibble() %>%
  rownames_to_column(var = "component") %>%
  ggplot() +
  aes(x=component, y=`percentage of variance`) +
  geom_col(width=.3) +
  ggtitle("CA Screeplot for attitudes to vaccine vers attitude to covid vaccine")
```

---

### Biplot

```{r, echo=FALSE}
fviz_ca_biplot(ca_xtab_general_covid)
```

---
exclude: true


```{r, echo=FALSE}
attitudes_vaccine <- c("Strongly in favor",
                       "Somewhat in favor",
                       "Somewhat not in favor",
                       "Not at all infavor")
```


```{r gender_vaccine, echo=FALSE}

### Factors associated with vaccination in general and Covid-19 refusals.

gender_vaccine <- rbind("Men"=c(11175 , 18599 , 6010 , 3240),
      "Women"=c(10145 , 23128, 8836 , 4722)) %>%
  as.table()

dimnames(gender_vaccine) <- list(Gender=rownames(gender_vaccine), Vaccine= attitudes_vaccine)
```

```{r age_vaccine, echo=FALSE}
age_vaccine <- rbind(
  "18-24" = c(2253, 4293 , 1423 , 721, 8690),
  "25-34" = c(2145, 4894 , 2065 , 1204 , 10308),
  "35-44" = c(3026, 7549 , 2668 , 1536, 14779),
  "45-54" = c(3462, 8343 , 3164 , 1689 , 16658),
  "55-64" = c(3589, 8051 , 2845 , 1404 , 15889),
  "+ 65"  = c(6845, 8597 , 2681 , 1408, 19531))[,1:4] %>%
  as.table

dimnames(age_vaccine) <- list(Age=rownames(age_vaccine), Vaccine= attitudes_vaccine)
```

```{r education_vaccine, echo=FALSE}
education_vaccine <- rbind(
  "No diploma" = c(831, 2025 , 794 , 769 , 4419),
  "Primary education" = c(1663 , 3373 , 1118 , 715 , 6869),
  "Vocational secondary" = c(3053 , 8225 , 3314 , 2320 , 16912),
  "High school" = c(4002, 9141, 3585 , 1877 , 18605),
  "High school + 2 to 4 years" = c(6233,  12585, 4374, 1745, 24937),
  "High school + 5 or more years" = c(5538, 6378, 1661, 536, 14113))[,1:4] %>%
  as.table()

dimnames(education_vaccine) <- list(Education=rownames(education_vaccine), Vaccine= attitudes_vaccine)
```

```{r csp_vaccine, echo=FALSE}
csp_vaccine <- rbind(
"Farmers" = c( 263 , 616 , 222 , 126 , 1227),
"Self-employed and entrepreneurs" = c( 1085 , 2265 , 950 , 562 , 4862 ),
"Senior executive professionals" = c( 7704 , 10088 , 2864 , 1105 , 21761 ),
"Middle executive professionals" = c( 4346 , 8889 , 3238, 1420, 17893 ),
"Employees" = c( 4052 , 11264 , 4517 , 2596,  22429 ),
"Manual workers" = c( 1551 , 4713 , 1910 , 1531 , 9705 ),
"Never worked and others" = c( 2319 , 3892 , 1145 , 622, 7978 ))[,1:4] %>%
  as.table()

dimnames(csp_vaccine) <- list(CSP=rownames(csp_vaccine), Vaccine= attitudes_vaccine)
```

```{r, echo=FALSE}
income_vaccine <- rbind(
"D1" = c( 1269 , 2748 , 1152 , 887 , 6056 ),
"D2-D3" = c( 1979 , 5012 , 2104 , 1480 , 10575 ),
"D4-D5" = c( 2528 , 6743 , 2667 , 1618 , 13556),
"D6-D7" = c( 3959 , 9222 , 3395 , 1692 , 18268 ),
"D8-D9" = c( 6503 , 12071 , 3892 , 1691 , 24157 ),
"D10" = c( 5082 , 5931 , 1636 , 594 , 13243 ))[,1:4] %>%
  as.table()

dimnames(income_vaccine) <- list(Income=rownames(income_vaccine), Vaccine= attitudes_vaccine)
```

```{r,  echo=FALSE}
ethnic_vaccine <- rbind(
"Mainstream population" = c( 17924 , 35201 , 12221 , 6117 , 71463 ),
"DOM or descendants of DOM native" = c( 194 , 457 , 191 , 154 , 996 ),
"Non-racialized second-generation immigrants" = c( 1132 , 2261 , 870 , 442 , 4705 ),
"Racialized second-generation immigrants" = c( 585 , 1315 , 560 , 414 , 2874 ),
"Non-racialized first-generation immigrants" = c( 840 , 1200 , 412 , 278 , 2730 ),
"Racialized first-generation immigrants" = c( 645 , 1293 , 592 , 557 , 3087 ))[,1:4] %>%
  as.table()

dimnames(ethnic_vaccine) <- list(Ethnos=rownames(ethnic_vaccine), Vaccine= attitudes_vaccine)
```


```{r, echo=FALSE}
child_vaccine <- rbind(
  "At least a child" = c(6987 , 17033 , 6200 , 3459 , 33679),
  "No child" = c(14333 , 24694 , 8646 , 4503 , 52176)
  )[,1:4] %>%
  as.table()

dimnames(child_vaccine) <- list(Children=rownames(child_vaccine), Vaccine= attitudes_vaccine)
```




```{r, echo=FALSE}
attitudes_covid_vaccine <- c("Yes probably", " Yes maybe", "Probably not", "Certainly not", "You do not know")
```

```{r gender_covid, echo=FALSE}

## Attitudes with covid vaccination

gender_covid <- rbind(
  "Men" = c( 16702 , 10865 , 3468 , 3707,  4282 , 39024 ),
  "Women" = c( 13122 , 12581 , 5992 , 7349 , 7787 , 46831)
  )[,1:5] %>%
  as.table()

dimnames(gender_covid) <-  list(Gender=rownames(gender_covid),
                                Covid_Vaccine=attitudes_covid_vaccine)

```

```{r age_covid, echo=FALSE}
age_covid <- rbind(
"18-24"  =  c(2892 , 2276 , 991 , 1440 , 1091 , 8690),
"25-34"  =  c(2428 , 2659 , 1623 , 2237 , 1361 , 10308),
"35-44"  =  c(3972 , 3971 , 2135 , 2550 , 2151 , 14779),
"45-54"  =  c(5168 , 4650 , 2029 , 2141 , 2670 , 16658),
"55-64"  =  c(5747 , 4694 , 1538 , 1534 , 2376 , 15889),
"+ 65"  =  c(9617 , 5196 , 1144 , 1154 , 2420 , 19531))[,1:5] %>%
as.table()

dimnames(age_covid) <- list(Age=rownames(age_covid), Covid_Vaccine=attitudes_covid_vaccine)
```

```{r education_covid, echo=FALSE}
education_covid <- rbind(
"No diploma" = c( 1273 , 1198 , 355 , 724 , 869 , 4419 ),
"Primary education" = c( 2507 , 1911 , 587 , 746 , 1118 , 6869 ),
"Vocational secondary" = c( 5052 , 4636 , 1619 , 2495 , 3110 , 16912 ),
"High school" = c( 5867 , 4952 , 2273 , 2865 , 2648 , 18605 ),
"High school + 2 to 4 years" = c( 8502 , 6963 , 3158 , 3148 , 3166 , 24937),
"High school + 5 or more years" = c( 6623 , 3786 , 1468 , 1078 , 1158 , 14113 ))[,1:5] %>%
as.table()

dimnames(education_covid) <- list(Education=rownames(education_covid), Covid_Vaccine=attitudes_covid_vaccine)
```

```{r csp_covid, echo=FALSE}
csp_covid <- rbind(
"Farmers" = c( 430 , 348 , 121 , 140 , 188 , 1227 ),
"Self-employed and entrepreneurs" = c( 1797 , 1229 , 529 , 652 , 655 , 4862 ),
"Senior executive professionals" = c( 10216 , 5861 , 2103 , 1626 , 1955 , 21761 ),
"Middle executive professionals" = c( 6065 , 5115 , 2075 , 2230 , 2408 , 17893 ),
"Employees" = c( 5871 , 6104 , 2798 , 3649 , 4007 , 22429 ),
"Manual workers" = c( 2632 , 2631 , 1015 , 1667 , 1760 , 9705 ),
"Never worked and others" = c( 2813 , 2158 , 819 , 1092 , 1096 , 7978 )) [,1:5] %>%
as.table()

dimnames(csp_covid) <- list(CSP=rownames(csp_covid), Covid_Vaccine=attitudes_covid_vaccine)
```

```{r income_covid, echo=FALSE}
income_covid <- rbind(
  "D1" = c( 1747 , 1501 , 635 , 1117 , 1056 , 6056 ),
  "D2-D3" = c( 2817 , 2714 , 1234 , 1940 , 1870 , 10575 ),
  "D4-D5" = c( 3765 , 3659 , 1655 , 2273 , 2204 , 13556 ),
  "D6-D7" = c( 5705 , 5183 , 2177 , 2499 , 2704 , 18268 ),
  "D8-D9" = c( 9198 , 6893 , 2633 , 2374 , 3059 , 24157 ),
  "D10" = c( 6592 , 3496 , 1126 , 853 , 1176 , 13243 ))[,1:5] %>%
  as.table()

dimnames(income_covid) <- list(Income=rownames(income_covid), Covid_Vaccine=attitudes_covid_vaccine)
```


```{r ethno_covid, echo=FALSE}
ethno_covid <- rbind(
  "Mainstream population" = c( 25375 , 19663 , 8094 , 8823 , 9508 , 71463 ),
  "DOM or descendants of DOM native" = c( 243 , 242 , 129 , 225 , 157 , 996 ),
  "Non-racialized second-generation immigrants" = c( 1591 , 1251 , 480 , 634 , 749 , 4705 ),
  "Racialized second-generation immigrants" = c( 751 , 667 , 327 , 588 , 541 , 2874 ),
  "Non-racialized first-generation immigrants" = c( 1041 , 750 , 189 , 331 , 419 , 2730 ),
  "Racialized first-generation immigrants" = c( 823 , 873 , 241 , 455 , 695 , 3087 )
  )[,1:5] %>%
   as.table()

dimnames(ethno_covid) <- list(Ethnos=rownames(ethno_covid), Covid_Vaccine=attitudes_covid_vaccine)
```


```{r children_covid, echo=FALSE}
children_covid <- rbind(
  "At least a child" = c( 9684 , 9220 , 4372 , 5292 , 5111 , 33679 ),
  "No child" = c( 20140 , 14226 , 5088 , 5764 , 6958 , 52176 ))[,1:5] %>%
  as.table()

dimnames(children_covid) <- list(Children=rownames(children_covid), Covid_Vaccine=attitudes_covid_vaccine)
```


```{r  afraid_covid, echo=FALSE}
afraid_covid <- rbind(
  "Yes" <- c(8842 , 5616 , 1477 , 1607 , 2626 , 20168 ),
  "No" <- c(20982 , 17830 , 7983 , 9449 , 9443 , 65687))[,1:5] %>%
  as.table()

dimnames(afraid_covid) <- list(
    Afraid=rownames(afraid_covid),
    Covid_Vaccine=attitudes_covid_vaccine)
```

```{r trust_government_covid, echo=FALSE}
trust_government_covid <- rbind(
  "Yes" = c( 19254 , 12656 , 3663 , 2764 , 5064 , 43401 ),
  "No" = c( 8777 , 8674 , 4910 , 7117 , 4647 , 34125 ),
  "You do not know" = c( 1793 , 2116 , 887 , 1175 , 2358 , 8329 ))[,1:5] %>%
  as.table(trust_government_covid)

dimnames(trust_government_covid) <- list(Trust=rownames(trust_government_covid),
  Covid_Vaccine=attitudes_covid_vaccine)
```



```{r comorbidities_covid, echo=FALSE}
comorbidities_covid <- rbind(
  "Yes" = c( 10456 , 6945 , 2288 , 2773 , 3542 , 26004 ),
  "No" = c( 19368 , 16501 , 7172 , 8283 , 8527 , 59851 ))[,1:5] %>%
  as.table()

dimnames(comorbidities_covid) <- list(Comorbidities=rownames(comorbidities_covid), Covid_Vaccine=attitudes_covid_vaccine)
```


```{r, echo=FALSE}
ls_tables_names  <- ls(pattern="_covid|_vaccine") 

mask <- map_lgl(ls_tables_names, ~ is.table(get(.)))

ls_tables <- ls_tables_names[mask] %>% 
  map(~ get(.))

names(ls_tables) <- ls_tables_names[mask]
```


```{r, echo=FALSE}
ca_analyses <- ls_tables %>% map( ~ CA(. , graph = FALSE))

# cas_covid <- ls_tables_covid %>% map(~ CA(., graph = FALSE))
```


---

### Age 

```{r age, out.width="50%", fig.show='hold', echo=FALSE}
fviz_ca(ca_analyses[[3]]) + ggtitle(names(ca_analyses)[[3]])
fviz_ca(ca_analyses[[2]]) + ggtitle(names(ca_analyses)[[2]])
```


---

### Age mosaicplots


```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(age_vaccine, shade = TRUE)
mosaicplot(age_covid, shade = TRUE)
```

---

### CSP 

```{r , out.width="50%", fig.show='hold', echo=FALSE}
fviz_ca(ca_analyses[[8]]) + ggtitle(names(ca_analyses)[[8]])
fviz_ca(ca_analyses[[7]]) + ggtitle(names(ca_analyses)[[7]])
```


---


### CSP mosaicplots


```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(csp_vaccine, shade = TRUE)
mosaicplot(csp_covid, shade = TRUE)
```

---

### Education

```{r, out.width="50%", fig.show='hold', echo=FALSE}
fviz_ca(ca_analyses[[10]]) + ggtitle(names(ca_analyses)[[10]])
fviz_ca(ca_analyses[[9]]) + ggtitle(names(ca_analyses)[[9]])
```

---

### Education mosaicplots

```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(education_vaccine, shade = TRUE)
mosaicplot(education_covid, shade = TRUE)
```
---

### Income

```{r, out.width="50%", fig.show='hold', echo=FALSE}
fviz_ca(ca_analyses[[16]]) + ggtitle(names(ca_analyses)[[16]])
fviz_ca(ca_analyses[[15]]) + ggtitle(names(ca_analyses)[[15]])
```

---

### Income mosaicplots

```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(income_vaccine, shade = TRUE)
mosaicplot(income_covid, shade = TRUE)
```


---

### Ethnicity


```{r, out.width="50%", fig.show='hold', echo=FALSE}
fviz_ca(ca_analyses[[11]]) + ggtitle(names(ca_analyses)[[11]])
fviz_ca(ca_analyses[[12]]) + ggtitle(names(ca_analyses)[[12]])
```


---

### Ethnicity mosaicplots

```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(ethnic_vaccine, shade = TRUE)
mosaicplot(ethno_covid, shade = TRUE)
```



---

### Gender


```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(gender_vaccine, shade = TRUE)
mosaicplot(gender_covid, shade = TRUE)

```


---

### Family 

```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(child_vaccine, shade = TRUE)
mosaicplot(children_covid, shade = TRUE)

```

---

### In government  we trust (or not)

```{r, out.width="50%", fig.show='hold', echo=FALSE}
mosaicplot(trust_government_covid, shade = TRUE)
mosaicplot(comorbidities_covid, shade = TRUE)
```


