---
title: "Unsupervised Learning"
author: "DSSP Team"
date: "Spring 2020"
output:
  html_document:
    css: hideOutput.css
    includes:
      in_header: hideOutput.script
---

```{r Knitr_Global_Options, include=FALSE}
library(knitr)

opts_chunk$set(warning = FALSE,
               message = FALSE,
               cache = TRUE,
               autodep = TRUE,
               tidy = FALSE)
```

# Principal Component Analysis

We will consider a _classical_ dataset __decathlon__. It contains the performance of several athletes during two decathlons in 2004 the Decastar and the Olympic Game. Each observation consists of

- the 10 raw performance in the 10 events,
- the ranking in the event,
- the total number of points,
- the name of the event.

We will mainly focus on the 10 first columns corresponding to the raw athletic performance but will also use the total number of points in some plots.

```{r Decathlon}
library("tidyverse")
library("FactoMineR")

data(decathlon)
colnames(decathlon)
glimpse(decathlon)
summary(decathlon)

old_theme <- theme_set(theme_light())
opts <- options()  # save old options

options(ggplot2.discrete.colour="viridis")
options(ggplot2.continuous.colour = "viridis")
```


## Pairwise analysis

1. Use `ggplot2` to plot the two first coordinates (__100m__ and __Long.jump__) adding the __Points__ information as a color as well as a regression line.

__Hint:__ You can use `ggrepel` to have a better text placement.

:::hiddensolution
```{r Decathlon_2D}
library("ggrepel")
decathlon2 <- decathlon %>%
  rownames_to_column()

decathlon2 %>%
  ggplot() +
  aes(x = `100m`, y = Long.jump) +
  aes(colour = Points, label=rowname) +
  geom_point(size = 5) +
  geom_smooth(se = FALSE) +
  geom_smooth(method="lm", se=FALSE) +
  geom_text_repel(box.padding = unit(0.75, "lines"))
```
:::


2. Do you think those values are correlated?

3. Compute the correlation matrix between the 10 first variables and display it use a heatmap.

:::hiddensolution
```{r Decathlon_Cor}
cor(decathlon[,1:10])
heatmap(abs(cor(decathlon[,1:10])),symm=T)
```
:::

4. Can you guess which are the most correlated variables and the least ones?

5. Confirm those findings by looking at the pairwise scatterplot.

:::hiddensolution
```{r Pairwise}
library(GGally)
ggpairs(decathlon2[,2:11])

ggpairs(decathlon2, columns = c(2,3,7,11))
```
:::

## 3D plot

We are now ready to use __plotly__ to visualize the dataset in 3d.

A 3d scatter plot can be obtained with the following code:

```{r Plot3D}
library(plotly)
p <- plot_ly() %>%
  add_markers(data = decathlon[, c(1:3, 12)],
              x = ~`100m`, y = ~Long.jump, z = ~Shot.put,
              color = ~Points,
              size=2,
              legendgroup='points',
              showscale=TRUE) %>%
  layout(scene = list(camera = list(projection = list(type = 'orthographic')),
                      xaxis = list(showspikes=FALSE),
                      yaxis = list(showspikes=FALSE),
                      zaxis = list(showspikes=FALSE)))
suppressWarnings(p)
```


6. Play with this interactive graphic to find he projection that appears to lose the less information?

7. This task is similar to the PCA where we look a the subspace that minimizes the error between the data and its projection. Draw the ellipse corresponding to the eigenvectors and the eigenvalues of the covariance matrix to verify this.

__Hint:__ Use the `mesh` layer of `plotly` with the `alphahull = 0` option and the following ellipse generating code:

```{r Plot3DEllipse}
sphere3d <- function() {
  theta <- seq(-pi/2, pi/2, by=0.1)
  ntheta <- length(theta)
  phi <- seq(0, 2*pi, by=0.1)
  nphi <- length(phi)
  theta_ <- as.vector(array(theta, dim = c(ntheta, nphi)))
  phi_ <- as.vector(aperm(array(phi, dim = c(nphi, ntheta)), c(2,1)))

  x <- cos(theta_) * cos(phi_)
  y <- cos(theta_) * sin(phi_)
  z <- sin(theta_)

  cbind(x=x, y=y, z=z)
}

ellipse3d <- function(mean, cov){
  cov_eig <- eigen(cov)
  cov_sqrt <- cov_eig$vectors %*% diag(sqrt(cov_eig$values)) %*% t(cov_eig$vectors)
  sphere <- sphere3d()
  setNames(
    as.data.frame(qnorm(.975) * sphere %*% cov_sqrt + aperm(array(mean, dim = rev(dim(sphere))), c(2, 1))),
    c('x', 'y', 'z'))
}
```

:::hiddensolution

```{r Plot3DEllipse2}
p <- plot_ly() %>%
  add_mesh(data = ellipse3d(colMeans(decathlon[,1:3]), cov(decathlon[,1:3])),
           x=~x, y=~y, z=~z,
           color = 8000,
           alphahull = 0, opacity=.1,
           showlegend=FALSE) %>%
  add_markers(data = decathlon[, c(1:3, 12)],
              x = ~`100m`, y = ~Long.jump, z = ~Shot.put,
              color = ~Points,
              size=2,
              legendgroup='points',
              showscale=TRUE) %>%
  layout(scene = list(camera = list(projection = list(type = 'orthographic')),
                      xaxis = list(showspikes=FALSE),
                      yaxis = list(showspikes=FALSE),
                      zaxis = list(showspikes=FALSE)))
suppressWarnings(p)
```
:::

8. Can you verify that the best subspace is defined by the direction of the eigenvectors with the largest eigenvalues?

## Scaling?

So far, we have use the raw performance. That is we compare duration with length as well as 100m time with 1500m time. Obviously, this is not a good idea. All those values should be measured in a common scale. The most classical technique is to _normalize_ by substracting its mean and dividing by the standard deviation. After doing thiw, we may repeat all the previous experiments with those rescaled data.

9. Make a copy of the decathlon dataframe and normalize the 10 first columns of this new data frame.

__Hint:__ Use `scale`

:::hiddensolution
```{r DecathlonR_2D}
decathlon  %>%
  rownames_to_column("Name") %>%     #<< tidyverse does not like rownames
  mutate(across(-c(Name, Rank, Points, Competition),
                ~ (.x -mean(.x))/sd(.x))) -> decathlonR2
```

:::

10. Plot the corresponding scatter plot and compare it with the one obtained with the raw data.

:::hiddensolution
```{r DecathlonR_Scat, out.width="100%", fig.show='hold'}
decathlonR2 %>%
  ggplot() +
  aes(x = `100m`, y = Long.jump) +
  aes(color = Points) +
  geom_text_repel(aes(label = Name),
                  box.padding = unit(0.75, "lines")) +
  geom_point(size = 5) +
  guides(color = FALSE) -> ggp_pair_comp

ggp_pair_comp
```

:::

11. What if one imposes _equal_ axis?

__Hint:__ Use `coord_fixed`.

:::hiddensolution

```{r Decathlon_Fixed}
ggp_pair_comp +
  coord_fixed() -> ggp_pair_comp

ggp_pair_comp
```

:::

12. Repeat the same experiment replacing the `100m` by the `1500m`.

:::hiddensolution

```{r Decathlon_1500m}
ggp_pair_comp +
  aes(x = `1500m`)
```
:::

13. What about the correlation matrix, the pairwise scatterplots and the 3d plot?

:::hiddensolution

```{r DecathlonR_Cor}
decathlonR2 %>%
  dplyr::select(-c(Name, Rank, Points, Competition)) %>%
  corrr::correlate(quiet = TRUE, method="pearson") %>%
  mutate(across(where(is_numeric), abs)) %>%
  corrr::rearrange() -> df_corr_decathlon


df_corr_decathlon %>%
  corrr::rplot() +
  coord_fixed() +
  ggtitle("Correlation plot for scaled Decathlon data") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
```

```{r RPairwise}
ggpairs(decathlonR2[,2:11])

ggpairs(decathlonR2, columns = c(2,3,7,11))
```

```{r RPlot3DEllipse}
p <- plot_ly() %>%
  add_mesh(data = ellipse3d(colMeans(decathlonR2[,2:4]), cov(decathlonR2[,2:4])),
           x=~x, y=~y, z=~z,
           color = 8000,
           alphahull = 0, opacity=.1, hoverinfo='none',
           showlegend=FALSE) %>%
  add_markers(data = decathlonR2[, 1 + c(1:3, 12)],
              x = ~`100m`, y = ~Long.jump, z = ~Shot.put,
              color = ~Points,
              size=2,
              legendgroup='points',
              showscale=TRUE) %>%
  layout(scene = list(camera = list(projection = list(type = 'orthographic')),
                      xaxis = list(showspikes=FALSE),
                      yaxis = list(showspikes=FALSE),
                      zaxis = list(showspikes=FALSE)))
suppressWarnings(p)
```
:::


## Principal Component Analysis

The package __FactoMineR__ contains a PCA function that we are going to use.

```{r PCA}
pacman::p_load("FactoMineR")

PCADecathlonR <- decathlonR2 %>%
  dplyr::select(-c(Name, Rank, Points, Competition)) %>%
  PCA(graph = FALSE,
      scale.unit = FALSE,
      ncp = 10)

decathlon_prcomp <- decathlonR2 %>%
  dplyr::select(-c(Name, Rank, Points, Competition)) %>%
  prcomp(x = ., center=FALSE, scale.=FALSE)

# decathlon_prcomp %>% summary()
# PCADecathlonR %>% summary()
```

See also [Tidy PCA through `prcomp`](https://rdrr.io/github/tidyverse/broom/man/tidy.prcomp.html)


`PCADecathlonR` contains all the information computed by the PCA (for instance the eigenvalues in the field `eig`, the SVD decomposition in `svd`). It also contains the coordinates of `DecathlonR` in the PCA basis.

14. Verify the the coordinates in the PCA basis can be obtained by a product between the coordinates in the old basis and the change of basis matrix $V$.

:::hiddensolution
```{r PCA_}
max(abs(PCADecathlonR$ind$coord -
        as.matrix(decathlonR2[2:11]) %*% PCADecathlonR$svd$V)
    )
```
:::

14. bis Verify the the coordinates in the PCA basis (attribute `x` in the output of `prcomp`) can be obtained by a product between the coordinates in the old basis and the change of basis matrix (attribute `rotation` in the output of `prcomp`).
:::hiddensolution
```{r PCA_2}
max(abs(decathlon_prcomp$x -
        as.matrix(decathlonR2[2:11]) %*% decathlon_prcomp$rotation)
    )
```
:::


15. Plot the points in the new coordinates

:::hiddensolution
```{r PCA_Plot}
PCADecathlonR$eig

ggplot(data= data.frame(X1 = PCADecathlonR$ind$coord[,1],
                        X2 = PCADecathlonR$ind$coord[,2],
                        Col = decathlon$Points),
       aes(x = X1, y = X2, color = Col)) +
  geom_point(size = 5) +
  geom_text_repel(label = row.names(decathlon)) +
  scale_x_continuous(expand = c(.15,0)) +
  scale_y_continuous(expand = c(.1,0)) +
  guides(color = FALSE) +
  coord_fixed() +
  labs(title= "Decathlon on factorial plane",
       x ="PC1 ")
```
:::


:::hiddensolution
```{r}
decathlon_prcomp %>%
  broom::augment() %>%
  tibble::add_column(Name = decathlonR2$Name,
                     Points= decathlonR2$Points) %>%
  ggplot() +
  aes(x= .fittedPC1, y=.fittedPC2) +
  aes(col=Points, label=Name) +
  geom_point(size = 5) +
  geom_text_repel() +     #<<
  scale_x_continuous(expand = c(.15,0)) +
  scale_y_continuous(expand = c(.1,0)) +
  guides(color = FALSE) +
  coord_fixed() +
  scale_color_viridis_c() +
  ggtitle("Decathlon on first two principal components")

```
:::

10. In statistics, one often looks at the correlation between the new axis and the old ones. This is already computed in __R__ by

```{r CoordsVar}
t(PCADecathlonR$var$coord)
```

Note that by construction the norm of the correlation vector for a given original axis is 1.

```{r CircleSum}
colSums(t(PCADecathlonR$var$coord)^2)
```

Plot the correlation of the original axes with the two first new components

:::hiddensolution
```{r CirclePlot}
ggplot(data.frame(X1 = PCADecathlonR$var$coord[,1],
                  X2 = PCADecathlonR$var$coord[,2]))+
  ggforce::geom_circle(aes(x0 = 0, y0 = 0, r = 1), color="grey") +
  geom_segment(aes(xend = X1, yend = X2),
               x = 0 , y = 0, arrow = grid::arrow()) +
  geom_text(aes(x = X1, y = X2),
            label = names(decathlonR2[2:11]),
            vjust = "outward", hjust = "outward") +
  coord_fixed() +
  geom_segment(x = 0, y = -1, xend = 0, yend = 1, linetype = "dashed", color = "grey") +
  geom_segment(x = -1, y = 0, xend = 1, yend = 0, linetype = "dashed", color = "grey")
```
:::


:::hiddenssolution
```{r}
decathlon_prcomp$rotation %>%
  as_tibble(rownames="Name") %>%
  ggplot() +
  aes(x=PC1, y=PC2, label=Name) +
  geom_segment(xend=0, yend=0, arrow = grid::arrow(ends = "first") ) +
  geom_text_repel() +
  coord_fixed() +
  xlim(-.7, .7) +
  ylim(-.7, .7) +
  stat_function(fun = ~  sqrt(.7^2 - .^2), alpha=.5, color="grey") +
  stat_function(fun = ~ - sqrt(.7^2 - .^2), alpha=.5, color="grey") +
  geom_segment(x = 0, y = -1, xend = 0, yend = 1, linetype = "dashed", color = "grey") +
  geom_segment(x = -1, y = 0, xend = 1, yend = 0, linetype = "dashed", color = "grey") +
  ggtitle("Decathlon correlation circle")


```
:::



:::hiddensolution
```{r}
biplot(decathlon_prcomp)
```
:::


17. Which variables are well captured? Can you interpret the new axes (the horizontal and the vertical ones)?
Can you explain why the long jump appears to be the opposite of the 100m?

__Hint:__ You may use the graph from `PCA` or from `factoextra`.

:::hiddensolution
```{r, out.width="50%", fig.show='hold'}
library(dplyr)
decathlonTime = decathlon[,1:10]

decathlonTime <- mutate(decathlonTime,
                        `100m` = 100/`100m`,
                        `400m` = 400/`400m`,
                        `110m.hurdle` = 110/`110m.hurdle`,
                        `1500m` = 1500/`1500m`)
rownames(decathlonTime)  <- rownames(decathlon)

PCAdecathlontime <- PCA(decathlonTime, graph = FALSE)

factoextra::fviz_pca_ind(PCAdecathlontime, repel = TRUE)
factoextra::fviz_pca_var(PCAdecathlontime, repel = TRUE)

# patchwork::wrap_plots(factoextra::fviz_pca_ind(PCAdecathlontime, repel = TRUE),
#                factoextra::fviz_pca_var(PCAdecathlontime, repel = TRUE),
#                ncol = 2)
```

:::

18. Plot the cumulative percentage of inertia. Do you think using 2 dimensions is enough here?

:::hiddensolution
```{r CumVariance}
ggplot(data = data.frame(dim = factor(1:length(PCADecathlonR$eig[,'cumulative percentage of variance'])),
                         cum = PCADecathlonR$eig[,'cumulative percentage of variance']),
       aes(x = dim, y = cum)) +
  geom_bar(stat = 'identity')
```
:::
